---
title: "Complete Analysis"
output: html_notebook
---

```{r message=F}
library(tidyverse)
library(ggpubr)
```

Data Import
===========

We have 60 sets of data in total. Each set represents a single trial of PM2.5,
PM10 readings for one group from the control and treatment groups. Each trial
was conducted using one of three sensor-box combinations (1a, 2b, or 3d).

Here, we define a function which will

1. Import a trial,
2. Add a column to documents its sensor-box and another for its group,
3. Divide the PM2.5 and PM10 readings, then for each of the two reading types:
4. Trim away all the readings before the first time the sensor desaturates, and
5. Set the time column as being the seconds since the first desaturated reading.

By desaturation, we mean the first reading which falls within the sensor's
specified range. For our model of PM sensor, this is defined as [0, 999.9]
micrograms per cubic metre.

```{r}
import <- function(filename) {
  data <- read_csv(filename, col_names=c('time', 'type', 'value'))
  data <- list(
    pm2.5=(data %>%
      subset(type=='pm2.5') %>%
      mutate(value=as.numeric(value)) %>%
      mutate(time=as.numeric(as.POSIXct(time)))),
    pm10=(data %>%
      subset(type=='pm10') %>%
      mutate(value=as.numeric(value)) %>%
      mutate(time=as.numeric(as.POSIXct(time))))
  )

  if (grepl('nothing', filename, fixed=T)) {
    data$group <- 'nothing'
  }
  else if (
    grepl('plant', filename, fixed=T) &
    !grepl('deadplant', filename, fixed=T) 
  ) {
    data$group <- 'plant'
  }
  else if (grepl('deadplant', filename, fixed=T)) {
    data$group <- 'deadplant'
  } else if (grepl('soil', filename, fixed=T)) {
    data$group <- 'soil'
  } else {
    stop(paste('import: Invaid group in', filename))
  }

  if(grepl('1[aA]', filename)) {
    data$sensorbox <- '1a'
  } else if (grepl('2[bB]', filename)) {
    data$sensorbox <- '2b'
  } else if (grepl('3[dD]', filename)) {
    data$sensorbox <- '3d'
  } else {
    stop(paste('import: Invalid sensor-box in', filename))
  }
  
  pm2.5_last_saturated <- last(which(data$pm2.5$value>=999.9))
  data$pm2.5 <- data$pm2.5[-(1:pm2.5_last_saturated),]
  pm10_last_saturated <- last(which(data$pm10$value>=999.9))
  data$pm10 <- data$pm10[-(1:pm10_last_saturated),]
  
  data$pm2.5$time <- data$pm2.5$time - data$pm2.5$time[1]
  data$pm10$time <- data$pm10$time - data$pm10$time[1]
  
  return(data)
}
```

Then, we import all our data by first building a vector of their filenames.

```{r}
filenames <- c(
  '19-octa-nothing-1a.csv',  '19-octa-nothing-2b.csv', '19-octa-nothing-3d.csv', 
  '20octa-nothing1a.csv',    '20octa-nothing2b.csv',   '20octa-nothing3d.csv', 
  '21octa-nothing-1a.csv',   '21octa-nothing-2b.csv',  '21octa-nothing-3d.csv', 
  '21octb-nothing-1a.csv',   '21octb-nothing-2b.csv',  '21octb-nothing-3d.csv', 
  '22octa-nothing-1a.csv',   '22octa-nothing-2b.csv',  '22octa-nothing-3d.csv', 
  '25octa-deadplant-1a.csv', '25octa-plant-3D.csv',    '25octa-soil-2b.csv', 
  '25octb-deadplant-1a.csv', '25octb-plant-3D.csv',    '25octb-soil-2b.csv', 
  '26octa-deadplant-1a.csv', '26octa-plant-3d.csv',    '26octa-soil-2b.csv', 
  '27octa-deadplant-1a.csv', '27octa-plant-3d.csv',    '27octa-soil-2b.csv', 
  '28octa-deadplant-1a.csv', '28octa-plant-3d.csv',    '28octa-soil-2b.csv', 
  '2nova-deadplant-2b.csv',  '2nova-plant-1a.csv',     '2nova-soil-3d.csv', 
  '2novb-deadplant-2b.csv',  '2novb-plant-1a.csv',     '2novb-soil-3d.csv', 
  '30octa-deadplant-2b.csv', '30octa-plant-1a.csv',    '30octa-soil-3d.csv', 
  '30octb-deadplant-2b.csv', '30octb-plant-1a.csv',    '30octb-soil-3d.csv', 
  '31octb-deadplant-2b.csv', '31octb-plant-1a.csv',    '31octb-soil-3d.csv', 
  '3nova-deadplant-3d.csv',  '3nova-plant-2b.csv',     '3nova-soil-1a.csv', 
  '4nova-deadplant-3d.csv',  '4nova-plant-2b.csv',     '4nova-soil-1a.csv', 
  '5nova-deadplant-3d.csv',  '5nova-plant-2b.csv',     '5nova-soil-1a.csv', 
  '6nova-deadplant-3d.csv',  '6nova-plant-2b.csv',     '6nova-soil-1a.csv', 
  '7nova-plant-2b.csv',  '7nova-deadplant-3d.csv', '7nova-soil-1a.csv'
) 
```

Followed by an `lapply` of the `import` function over the filenames vector to
obtain a list of lists. Each nested list is a trial.

```{r message=F}
data <- lapply(filenames, function(fn) {
  import(paste('05-analysis-proper/', fn, sep=''))})
```

Sensor Calibration
==================

But, each sensor labelled 1 to 3 are not calibrated with each other. Given the
same 'air', they give different readings. In order to calibrate these sensors,
we use a set of time-synchronised readings between the 3 sensors, where

- Each sensor is placed in the same box and reads during the same session, and
- Each sensor reads within 5s of each other.

An arbitrary sensor, the one labelled 3, is selected as the baseline for
calibration.

First, load the readings for the individual sensors. Then, combine them into
one tibble. Note that each sensor's time stamp is synchronised because they are
given by the same raspberry pi zero computer.

```{r message=F}
import_cdata <- function(filename) {
  data <- filename %>%
    read_csv(col_names=c('time', 'type', 'value')) %>%
    subset(type != 'comment') %>%
    mutate(value=as.numeric(value)) %>%
    mutate(time=as.numeric(as.POSIXct(time)))
}

# The three sensors were set-up such that they all began at the same time, +- 1s.
sensor1 <- import_cdata('03-sensor-calibration/16octc-nothing-1bigbox.csv')
sensor2 <- import_cdata('03-sensor-calibration/16octc-nothing-2bigbox.csv')
sensor3 <- import_cdata('03-sensor-calibration/16octc-nothing-3bigbox.csv')

# Some times, the an interval (supposedly per-minute) fails to read, and a row is missing.
# So, we want to make sure that each row index of sensor 1/2 aligns with sensor 3.
grouped_pm2.5 <- tibble(s1=double(), s2=double(), s3=double())
sensor1_pm2.5 <- subset(sensor1, type=='pm2.5')
sensor2_pm2.5 <- subset(sensor2, type=='pm2.5')
sensor3_pm2.5 <- subset(sensor3, type=='pm2.5')
# Also, trim the pre-desaturation data. The 999.9 are outliers with high leverage, and the
# data leading up to saturation is high variance.
last_saturated <- max(
  last(which(sensor1_pm2.5$value >= 999.9)),
  last(which(sensor2_pm2.5$value >= 999.9)),
  last(which(sensor3_pm2.5$value >= 999.9))
)
sensor1_pm2.5 <- sensor1_pm2.5[-(1:last_saturated),]
sensor2_pm2.5 <- sensor2_pm2.5[-(1:last_saturated),]
sensor3_pm2.5 <- sensor3_pm2.5[-(1:last_saturated),]
for (i in seq(1, nrow(sensor3_pm2.5))) {
  s3 <- sensor3_pm2.5[i,]
  s1 <- sensor1_pm2.5 %>% mutate(time=time-s3[1,]$time) %>% subset(abs(time) <= 5)
  s2 <- sensor2_pm2.5 %>% mutate(time=time-s3[1,]$time) %>% subset(abs(time) <= 5)
  if (nrow(s1) == 0 & nrow(s2) == 0) {
    next
  } else if (nrow(s1) == 0) {
    grouped_pm2.5 <- bind_rows(grouped_pm2.5, list(s1=NA, s2=s2[1,]$value, s3=s3[1,]$value))
  } else if (nrow(s2) == 0) {
    grouped_pm2.5 <- bind_rows(grouped_pm2.5, list(s1=s1[1,]$value, s2=NA, s3=s3[1,]$value))
  } else {
    grouped_pm2.5 <- bind_rows(grouped_pm2.5, list(s1=s1[1,]$value, s2=s2[1,]$value, s3=s3[1,]$value))
  }
}
# Repeat for PM10...
grouped_pm10 <- tibble(s1=double(), s2=double(), s3=double())
sensor1_pm10 <- subset(sensor1, type=='pm10')
sensor2_pm10 <- subset(sensor2, type=='pm10')
sensor3_pm10 <- subset(sensor3, type=='pm10')
last_saturated <- max(
  last(which(sensor1_pm10$value >= 999.9)),
  last(which(sensor2_pm10$value >= 999.9)),
  last(which(sensor3_pm10$value >= 999.9))
)
sensor1_pm10 <- sensor1_pm10[-(1:last_saturated),]
sensor2_pm10 <- sensor2_pm10[-(1:last_saturated),]
sensor3_pm10 <- sensor3_pm10[-(1:last_saturated),]
for (i in seq(1, nrow(sensor3_pm10))) {
  s3 <- sensor3_pm10[i,]
  s1 <- sensor1_pm10 %>% mutate(time=time-s3[1,]$time) %>% subset(abs(time) <= 5)
  s2 <- sensor2_pm10 %>% mutate(time=time-s3[1,]$time) %>% subset(abs(time) <= 5)
  if (nrow(s1) == 0 & nrow(s2) == 0) {
    next
  } else if (nrow(s1) == 0) {
    grouped_pm10 <- bind_rows(grouped_pm10, list(s1=NA, s2=s2[1,]$value, s3=s3[1,]$value))
  } else if (nrow(s2) == 0) {
    grouped_pm10 <- bind_rows(grouped_pm10, list(s1=s1[1,]$value, s2=NA, s3=s3[1,]$value))
  } else {
    grouped_pm10 <- bind_rows(grouped_pm10, list(s1=s1[1,]$value, s2=s2[1,]$value, s3=s3[1,]$value))
  }
}

data_calibration <- list(pm2.5=grouped_pm2.5, pm10=grouped_pm10)
```

A quick graphical check to show that the sensors are indeed different:

```{r}
ggplot(data_calibration$pm2.5) +
  geom_point(mapping=aes(x=s3, y=s1), size=.2, colour='blue') +
  geom_point(mapping=aes(x=s2, y=s1), size=.2, colour='red') +
  geom_abline(linetype='longdash')
ggplot(data_calibration$pm10) +
  geom_point(mapping=aes(x=s3, y=s1), size=.2, colour='blue') +
  geom_point(mapping=aes(x=s2, y=s1), size=.2, colour='red') +
  geom_abline(linetype='longdash')
```

Now, the calibration via regression.

```{r}
calibration <- list(
  pm2.5=list(
    s1=lm(s1~s3+I(s3^2), data=data_calibration$pm2.5),
    s2=lm(s2~s3+I(s3^2), data=data_calibration$pm2.5)
  ),
  pm10=list(
    s1=lm(s1~s3+I(s3^2), data=data_calibration$pm10),
    s2=lm(s2~s3+I(s3^2), data=data_calibration$pm10)
  )
)

summary(calibration$pm2.5$s1)
summary(calibration$pm2.5$s2)
summary(calibration$pm10$s1)
summary(calibration$pm10$s2)
```

Finally, package it all into a function.

```{r}
calibrate_one <- function(pm, sensorbox, x) {
  if (is.na(x)) return(NA)
  if (pm == 'pm2.5' & sensorbox == '1a') {
    lm_obj <- calibration$pm2.5$s1
  } else if (pm == 'pm2.5' & sensorbox == '2b') {
    lm_obj <- calibration$pm2.5$s2
  } else if (pm == 'pm10' & sensorbox == '1a') {
    lm_obj <- calibration$pm10$s1
  } else if (pm == 'pm10' & sensorbox == '2b') {
    lm_obj <- calibration$pm10$s2
  } else {
    stop(paste('Invalid pm/sensorbox: ', pm, sensorbox))
  }
  coeff <- lm_obj$coefficients - c(x, 0, 0)
  roots <- Re(polyroot(coeff))
  root <- roots[roots>=0]
  if (length(root) == 1) return(root)
  return(NA)  # would use NULL, but can't have NA in the middle of a vector
}

calibrate <- function(pm, sensorbox, xs) {
  return(unlist(Map(
    function(x) calibrate_one(pm, sensorbox, x), xs)))
}
```

Some closing graphical sanity checks:

```{r}
ggplot(data_calibration$pm2.5 %>% mutate(s1.fix=calibrate('pm2.5', '1a', s1))) +
  geom_point(mapping=aes(x=s3, y=s1), colour='red', size=.2) +
  geom_point(mapping=aes(x=s3, y=s1.fix), colour='darkgreen', size=.2) +
  geom_abline()
ggplot(data_calibration$pm2.5 %>% mutate(s2.fix=calibrate('pm2.5', '2b', s2))) +
  geom_point(mapping=aes(x=s3, y=s2), colour='red', size=.2) +
  geom_point(mapping=aes(x=s3, y=s2.fix), colour='darkgreen', size=.2) +
  geom_abline()
ggplot(data_calibration$pm10 %>% mutate(s1.fix=calibrate('pm10', '1a', s1))) +
  geom_point(mapping=aes(x=s3, y=s1), colour='red', size=.2) +
  geom_point(mapping=aes(x=s3, y=s1.fix), colour='darkgreen', size=.2) +
  geom_abline()
ggplot(data_calibration$pm10 %>% mutate(s2.fix=calibrate('pm10', '2b', s2))) +
  geom_point(mapping=aes(x=s3, y=s2), colour='red', size=.2) +
  geom_point(mapping=aes(x=s3, y=s2.fix), colour='darkgreen', size=.2) +
  geom_abline()
```

PM2.5 Box Plot & Multilinear Regression
=======================================

First, calibrate the data.

```{r}
data_calibrated <- data %>%
  lapply(function(trial) {
    if (trial$sensorbox == '1a') {
      trial$pm2.5$value <- calibrate('pm2.5', '1a', trial$pm2.5$value)
      trial$pm10$value <- calibrate('pm10', '1a', trial$pm10$value)
    } else if (trial$sensorbox == '2b') {
      trial$pm2.5$value <- calibrate('pm2.5', '2b', trial$pm2.5$value)
      trial$pm10$value <- calibrate('pm10', '2b', trial$pm10$value)
    } else if (trial$sensorbox == '3d') {
      # noop
    } else {
      stop(paste('Invalid sensorbox', trial$sensorbox))
    }
    return(trial)
  })
```
```{r, include=F}
# Quick sanity check. Toggle include=T to show.
lapply(data_calibrated, function(trial) {
  ggplot(trial$pm2.5) + geom_line(mapping=aes(x=time, y=value)) + xlim(NA, 3*60*60)
  ggplot(trial$pm10) + geom_line(mapping=aes(x=time, y=value)) + xlim(NA, 3*60*60)
})
```

Now, create a new tibble which records the PM decrease 2 hours after the first
sub-900 micrograms per cubic metre reading.

```{r}
pm2.5_aggregated <- lapply(data_calibrated, function(trial) {
  last_super900 <- last(which(trial$pm2.5$value >= 900))
  valid_readings <- trial$pm2.5[-(1:last_super900),]
  initial_reading <- valid_readings[1,]$value
  readings_aft_2hrs <- valid_readings %>%
    mutate(time=time-(2*60*60)-valid_readings[1,]$time) %>%
    subset(abs(time) <= 5)
  last_reading <- readings_aft_2hrs[1,]$value
  list(sensorbox=trial$sensorbox, group=trial$group, decrease=initial_reading-last_reading)
})

pm2.5_decrease <- tibble(
  sensorbox=sapply(pm2.5_aggregated, function(xs) xs$sensorbox),
  group=sapply(pm2.5_aggregated, function(xs) xs$group),
  decrease=sapply(pm2.5_aggregated, function(xs) xs$decrease)
)

pm2.5_decrease$has_soil <- ifelse(pm2.5_decrease$group %in% c('soil', 'deadplant', 'plant'), 1, 0)
pm2.5_decrease$has_sa <- ifelse(pm2.5_decrease$group %in% c('deadplant', 'plant'), 1, 0)
pm2.5_decrease$has_life <- ifelse(pm2.5_decrease$group == 'plant', 1, 0)
```

The multilinear regression:

```{r}
pm2.5_decrease.lm <- lm(decrease~has_soil + has_sa + has_life + sensorbox, data=pm2.5_decrease)
summary(pm2.5_decrease.lm)
```

The boxplot:

```{r}
groups <- factor(pm2.5_decrease$group, levels=c('nothing', 'soil', 'deadplant', 'plant'))
pm2.5_decrease$group <- groups

ggboxplot(pm2.5_decrease, x='group', y='decrease', color='sensorbox') +
    ylab(expression(atop("PM"[2.5]*" decrease ("*mu*"g m"^-3*")", ""))) +
    xlab("\nControl/Treatment Group") +
    guides(color=guide_legend(title='Box')) +
    scale_x_discrete(labels=c(
        'nothing'='Nothing',
        'soil'='Substrate Only',
        'deadplant'=expression('Dried Specimen'),
        'plant'=expression('Live Specimen'))) +
    scale_color_discrete(labels=c('Box 1', 'Box 2', 'Box 3')) +
    theme_grey()
ggsave('boxplot-pm25.eps', width=9, height=5)
```

PM10 Boxplot & Multilinear Regression
=====================================

```{r}
pm10_aggregated <- lapply(data_calibrated, function(trial) {
  last_super900 <- last(which(trial$pm10$value >= 900))
  valid_readings <- trial$pm10[-(1:last_super900),]
  initial_reading <- valid_readings[1,]$value
  readings_aft_2hrs <- valid_readings %>%
    mutate(time=time-(2*60*60)-valid_readings[1,]$time) %>%
    subset(abs(time) <= 5)
  last_reading <- readings_aft_2hrs[1,]$value
  list(sensorbox=trial$sensorbox, group=trial$group, decrease=initial_reading-last_reading)
})

pm10_decrease <- tibble(
  sensorbox=sapply(pm10_aggregated, function(xs) xs$sensorbox),
  group=sapply(pm10_aggregated, function(xs) xs$group),
  decrease=sapply(pm10_aggregated, function(xs) xs$decrease)
)

pm10_decrease$has_soil <- ifelse(pm10_decrease$group %in% c('soil', 'deadplant', 'plant'), 1, 0)
pm10_decrease$has_sa <- ifelse(pm10_decrease$group %in% c('deadplant', 'plant'), 1, 0)
pm10_decrease$has_life <- ifelse(pm10_decrease$group == 'plant', 1, 0)
```

The multilinear regression:

```{r}
pm10_decrease.lm <- lm(decrease~has_soil + has_sa + has_life + sensorbox, data=pm10_decrease)
summary(pm10_decrease.lm)
```

The boxplot:

```{r}
groups <- factor(pm10_decrease$group, levels=c('nothing', 'soil', 'deadplant', 'plant'))
pm10_decrease$group <- groups

ggboxplot(pm10_decrease, x='group', y='decrease', color='sensorbox') +
    ylab(expression(atop("PM"[10]*" decrease ("*mu*"g m"^-3*")", ""))) +
    xlab("\nControl/Treatment Group") +
    guides(color=guide_legend(title='Box')) +
    scale_x_discrete(labels=c(
        'nothing'='Nothing',
        'soil'='Substrate Only',
        'deadplant'=expression('Dried Specimen'),
        'plant'=expression('Live Specimen'))) +
        scale_color_discrete(labels=c('Box 1', 'Box 2', 'Box 3')) +
    theme_grey()
ggsave('boxplot-pm10.eps', width=9, height=5)
```
